---
title: "April–June 2018 Forecast"
---

## {- .tabset .tabset-fade .tabset-pills}

```{r setup, include=FALSE}
source("./R/setup.R")
```

### <img src="icons/enso.svg" width="35px"/> El Niño/La Niña {-}

Pacific sea surface temperatures have remained below average, maintaining La Niña conditions in the ocean and lower atmosphere. This typically results in wetter and cooler conditions during Montana’s winters and spring. Early in the spring the La Niña conditions are expected to contribute to below-average temperatures and above-median precipitation for portions of Montana. Most models agree that the La Niña will transition to ENSO neutral in late April to May. When ENSO-neutral occurs, ocean temperatures and atmospheric conditions are near the long-term average and there is often less predictability in the summer three month outlook. 

#### Temperature {-}
```{r future-seasonal-enso-temp, fig.retina = 2, fig.align = "center", out.width = '100%', collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}


enso_temp <- mtd_plot_enso(months = 4:6,
                                element = "tmpc",
                                agg_fun = mean)
 
enso_temp$map +
  add_climate_divisions()
```

#### Precipitation {-}
```{r future-seasonal-enso-prcp, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

enso_prcp <- mtd_plot_enso(months = 4:6,
                                element = "pcpn",
                                agg_fun = sum)
 
enso_prcp$map +
  add_climate_divisions()
```


### <img src="icons/temperature.svg" width="35px"/> Temperature {-}

Because of the weakening La Niña, the spring outlook for temperature is cooler than normal across eastern, central, and northwestern portions of the state. Equal chances of above, below, or normal spring temperatures are expected for the southwestern portions of the state. If the cooler than normal temperatures are realized for the majority of the state they will contribute to a delay in the onset of warmer soil temperatures, the growing season, and snowmelt. 

```{r future-seasonal-temp, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

climdiv_tmpc <- climdiv_summary(months = 4:6,
                                year = 2017,
                                element = "tmpc",
                                agg_fun = mean)


noaa_seasonal <- mtd_plot_noaa_seasonal_forecast(date = "2018-04-01",
                                                 element = "temp",
                                                 data_out = "../data/NOAA_seasonal")

noaa_seasonal$map +
  add_climate_divisions() +
  geom_label(data = climdiv_tmpc$normals %>%
               left_join(mt_climate_divisions_simple %>%
                           dplyr::mutate(Centroid_x = mt_climate_divisions_simple %>%
                                           sf::st_centroid() %>%
                                           sf::st_coordinates() %>%
                                           tibble::as_tibble() %$%
                                           X,
                                         Centroid_y = mt_climate_divisions_simple %>%
                                           sf::st_centroid() %>%
                                           sf::st_coordinates() %>%
                                           tibble::as_tibble() %$%
                                           Y),
                         .,
                         by = "Division code"),
             aes(x = Centroid_x,
                 y = Centroid_y,
                 label = stringr::str_c("1981-2010 normal: \n",
                                        Normal %>%
                                          round(digits = 1) %>%
                                          paste0(" ºF"))),
             alpha = 1,
             size = 2.25)

```

### <img src="icons/precipitation.svg" width="35px"/> Precipitation {-}
The weakening La Niña still plays a part for expected above normal precipitation in northeastern to northcentral Montana. Here the polar jetstream is expected to continue bringing moisture from the north and is welcome for portions of northeast Montana that still have moisture deficits from last summer’s drought. Across the rest of the state spanning from southeastern to western Montana, precipitation amounts are less certain and equal chances of above, below or normal precipitation are forecast.

```{r future-seasonal-prcp, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

climdiv_pcpn <- climdiv_summary(months = 4:6,
                                year = 2017,
                                element = "pcpn",
                                agg_fun = sum)

noaa_seasonal <- mtd_plot_noaa_seasonal_forecast(date = "2018-04-01",
                                                 element = "prcp",
                                                 data_out = "../data/NOAA_seasonal")

noaa_seasonal$map +
  add_climate_divisions() +
  geom_label(data = climdiv_pcpn$normals %>%
               left_join(mt_climate_divisions_simple %>%
                           dplyr::mutate(Centroid_x = mt_climate_divisions_simple %>%
                                           sf::st_centroid() %>%
                                           sf::st_coordinates() %>%
                                           tibble::as_tibble() %$%
                                           X,
                                         Centroid_y = mt_climate_divisions_simple %>%
                                           sf::st_centroid() %>%
                                           sf::st_coordinates() %>%
                                           tibble::as_tibble() %$%
                                           Y),
                         .,
                         by = "Division code"),
             aes(x = Centroid_x,
                 y = Centroid_y,
                 label = stringr::str_c("Normal: \n",
                                        Normal %>%
                                          round(digits = 1) %>%
                                          paste0(" in."))),
             alpha = 1,
             size = 2.25)

```

### <img src="icons/drought.svg" width="35px"/> Drought {-}

```{r future-seasonal-drought, fig.retina = 2, fig.align = "center", out.width = '100%', collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

future_seasonal_drought <- mtd_plot_noaa_drought_outlook(date = "2018-04-01",
                              data_out = "../data/NOAA_droughtlook")

future_seasonal_drought$map
```

