---
title: "January--March Mid-century Outlook"
editor_options: 
  chunk_output_type: console
---

## {- .tabset .tabset-fade .tabset-pills}

```{r setup, include=FALSE}
source("./R/setup.R")
```

Understanding how climate change will impact agriculture in the future can be difficult. It is often helpful to compare future climate projections to the conditions that we are currently experiencing. Below, we compare mid-century temperature and precipitation projections for January through March to our current normal (1981--2010). In short, mid-century winter and spring will look a lot different than what we are currently experiencing in terms of increasing temperatures and reduced snowpack.

### <img src="icons/temperature.svg" width="35px"/> Temperature {-}

By mid-century (2040--2069) average temperatures from January through March are expected to be 4--6 degrees warmer than current temperatures. Generally, temperature is projected to increase more in eastern and central Montana relative to the northwest portions of the state. Increased temperatures will result in an earlier start to the growing season. In addition, snowmelt will occur earlier and a larger proportion of our February and March precipitation will be in the form of rain as opposed to snow.

```{r midcentury, collapse=TRUE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}

macav2_midcentury <- 
  mtd_get_macav2_monthly_midcentury_summary(months = 1:3,
                                            data_out = "../data")

```


```{r midcentury-seasonal-temp-grid, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc")
```


```{r midcentury-seasonal-temp-grid-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   use_normals = TRUE)
```


```{r midcentury-seasonal-temp-counties, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median)
```


```{r midcenturyt-seasonal-temp-counties-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median,
                   use_normals = TRUE)
```

### <img src="icons/precipitation.svg" width="35px"/> Precipitation {-}
By mid-century (2040--2069) average total precipitation from January through March is expected to be similar or slightly higher than our current precipitation. Because of increasing temperatures, a larger portion of total precipitation will be in the form of rain as opposed to snow.

```{r midcentury-seasonal-prcp-grid, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn")
```


```{r midcentury-seasonal-prcp-grid-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   use_normals = TRUE)
```


```{r midcentury-seasonal-prcp-counties, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median)
```


```{r midcenturyt-seasonal-prcp-counties-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median,
                   use_normals = TRUE)
```

### <img src="icons/swe.svg" width="35px"/> Snowpack {-}
Due to increasing temperatures a larger proportion of Montana’s total winter precipitation will in the form of rain as opposed to snow. Increased temperatures will result in earlier snowmelt and a significant decline in our April 1<sup>st</sup> snowpack. By mid-century (2040--2069), it is expected that April 1<sup>st</sup> SWE in western and southcentral Montana will decrease to 25–67% of our current normal values. Progressing further east, snowpack in central to eastern Montana is expected to decrease to between 1--10% of its current normal April 1<sup>st</sup> snow water equivalent. By mid-century, very little snowpack will persist into April in central to eastern Montana.

```{r midcentury-swe, collapse=TRUE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}

midcentury_swe <- mtd_plot_midcentury_swe_basins(month = 4,
                                                 data_out = "../data/BCSD_mon_VIC")

midcentury_swe$map
```
