---
title: "April–June Mid-century Outlook"
editor_options: 
  chunk_output_type: console
---

## {- .tabset .tabset-fade .tabset-pills}

```{r setup, include=FALSE}
source("./R/setup.R")
```

Understanding how climate change will impact agriculture yield in the future can be difficult. It is often easier to contemplate what future projections of climate will look like when comparing them to a time period that we are currently experiencing. Below, we compared mid-century temperature and precipitation projections for January to March to our current 1981–2010 normal. In short, future mid-century winters and spring will look a lot different than what we are currently experiencing in terms of increasing temperatures and reduced snowpack.

### <img src="icons/temperature.svg" width="35px"/> Temperature {-}

January to March average temperatures in the period from 2040--2069 are expected to be from 4 to 6 degrees warmer relative to our current temperatures. Generally, the amount of temperature increase is estimated to be greater in eastern and central Montana relative to northwest portions of the state. Increased temperatures will result in an earlier start to the growing season. In addition snowmelt will occur earlier and a larger proportion of our February and March precipitation will be in the form of rain and not snow.

```{r midcentury, collapse=TRUE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}

macav2_midcentury <- 
  mtd_get_macav2_monthly_midcentury_summary(months = 1:3,
                                            data_out = "../data")

```


```{r midcentury-seasonal-temp-grid, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc")
```


```{r midcentury-seasonal-temp-grid-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   use_normals = TRUE)
```


```{r midcentury-seasonal-temp-counties, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median)
```


```{r midcenturyt-seasonal-temp-counties-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "tmpc",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median,
                   use_normals = TRUE)
```

### <img src="icons/precipitation.svg" width="35px"/> Precipitation {-}
January to March average total precipitation in the period from 2040--2069 is expected to be similar to slightly higher than our current precipitation amounts. Because of increasing temperatures (above), a larger portion of this total precipitation will occur as rainfall.

```{r midcentury-seasonal-prcp-grid, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn")
```


```{r midcentury-seasonal-prcp-grid-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   use_normals = TRUE)
```


```{r midcentury-seasonal-prcp-counties, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median)
```


```{r midcenturyt-seasonal-prcp-counties-normals, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
scratch <- macav2_midcentury %>%
  mtd_plot_macav2_monthly_midcentury_summary(element = "pcpn",
                   agg_sf = mt_counties_simple,
                   agg_sf_fun = median,
                   use_normals = TRUE)
```

### <img src="icons/swe.svg" width="35px"/> Snowpack {-}
Due to increasing temperatures a larger proportion of Montana’s total spring precipitation will occur as rainfall. Increased temperatures will result in earlier snowmelt and a significant decline in our April 1st snowpack. By the time period of 2040--2069, it is expected that April 1st snow water in western to southcentral portions of the state will decrease to 25--67% of our current normal values. Progressing further east, the plains snowpack in central to eastern Montana is expected to decrease to between 10--1% of its current normal April 1st snow water equivalent. By mid-century very little snowpack will persist into April in central to eastern Montana

```{r midcentury-swe, collapse=TRUE, results='hide', message=FALSE, warning=FALSE, cache=TRUE}

midcentury_swe <- mtd_plot_midcentury_swe_basins(month = 4,
                                                 data_out = "../data/BCSD_mon_VIC")

midcentury_swe$map
```
