---
title: "January--March, 2018: In review"
---

## {- .tabset .tabset-fade .tabset-pills}

```{r setup, include=FALSE}
source("./R/setup.R")
```

La Nina didn't dissappoint! Reflecting on last winter...

### <img src="./icons/temperature.svg" width="35px"/> Temperature {-}
```{r past-seasonal-temp, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

past_seasonal_temp <- mtd_plot_climdiv(months = 1:2,
                                 year = 2018,
                                 element = "tmpc",
                                 agg_fun = mean)

past_seasonal_temp$map

```

* In January--February, 2018, the average temperature statewide was `r round(past_seasonal_temp$data$state_rank$Value, digits = 1)` ºF, `r abs(round(past_seasonal_temp$data$state_rank$Deviation, digits = 1))` degrees lower than the 1981--2010 normal value.
* January--February, 2018 was the `r scales::ordinal(past_seasonal_temp$data$state_rank$Rank)` coldest winter since record keeping began in 1895.

### <img src="./icons/precipitation.svg" width="35px"/> Precipitation {-}
```{r past-seasonal-prcp, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

past_seasonal_prcp <- mtd_plot_climdiv(months = 1:2,
                                 year = 2018,
                                 element = "pcpn",
                                 agg_fun = sum)

past_seasonal_prcp$map

```

* In January--February, 2018, the average total precipitation statewide was `r round(past_seasonal_prcp$data$state_rank$Value, digits = 1)` inches, `r round(abs(past_seasonal_prcp$data$state_rank$Deviation), digits = 1)` inches lower than the 1981--2010 normal value.
* January--February, 2018 was the `r scales::ordinal(past_seasonal_prcp$data$state_rank$Rank)` driest winter since record keeping began in 1895.

### <img src="./icons/swe.svg" width="35px"/> Snowpack {-}

```{r past-seasonal-swe, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}
past_seasonal_swe <- mtd_plot_swe_basins(date="2018-04-01")

past_seasonal_swe$map

```

#### Big Story: Blackfeet Blizzard {-}


### <img src="./icons/soil_saturation.svg" width="35px"/> Soil Moisture {-}

```{r past-seasonal-smap, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

past_seasonal_smap <- mtd_plot_smap(date = "2018-04-01",
                                    data_out = "../data/SMAP")

past_seasonal_smap$map

```

#### Snow and soil moisture recharge {-}
"Recharge rates" ...

### ![drought icon](./icons/drought.svg){width="35px"} Drought {-}
```{r past-seasonal-drought, collapse=TRUE, results='hold', message=FALSE, warning=FALSE, cache=TRUE}

past_seasonal_drought <- mtd_plot_usdm(date = "2018-04-01",
                                       data_out = "../data/USDM")

past_seasonal_drought$map

```

Not out of it yet in the northeast.

